---
title: "Predicting Covid-19 EMS Incidents from Daily Hospitalization Trends"
author: "Ngoc Mai Tran, Evdokia Nikolova, David Kulpanowski, Yangxinyu Xie, Joshua Ong"
output: github_document
---

## Load Incident Data

```{r load}
setwd("~/Desktop/ems/data/final/")
rm(list = ls())
Incidents <- read.csv("Incidents2019_2020.csv")
ts.df <- read.csv("ts_2019_2020.csv")
```

## Basic Data Description

Call Disposition

```{r Basic Statistics Call Disposition}
library(ggplot2)
table_call <- summary(Incidents$Call_Disposition)
table_call <- table_call[table_call > 0]
s <- sum(table_call)
table_call <- data.frame(table_call[table_call > 5000])
colnames(table_call) <- "Count"
table_call$Percentage <- format(table_call$Count/s * 100, digits = 3)
table_call$Call_Disposition = row.names(table_call)

print("Proportion of the 6 major hospitals")

sum(table_call[table_call$Call_Disposition %in% c("Dell Seton Med Ctr", "South Austin Hospital", "Saint Davids Med Ctr", "North Austin Hospital", "Seton Med Ctr", "Seton Northwest"),]$Count)/s

colors <- c("Refusal" = "#575655", "Other" = "#575655", "Call Cancelled" = "#575655", "Referred" = "#575655",
            "No Patient" = "#575655", "Dell Seton Med Ctr" = "#029ac4", "North Austin Hospital" = "#029ac4", "Saint Davids Med Ctr" = "#029ac4", "Seton Med Ctr" = "#029ac4", "Seton Northwest" = "#029ac4", "South Austin Hospital" = "#029ac4")
g1 <- ggplot(data=table_call, aes(x=reorder(Call_Disposition, Count),y=Percentage, fill = Call_Disposition)) +
    scale_fill_manual(limits = c("Dell Seton Med Ctr", "South Austin Hospital", "Saint Davids Med Ctr", "North Austin Hospital", "Seton Med Ctr", "Seton Northwest", "Refusal", "Call Cancelled", "Referred", "No Patient", "Other"), values = colors) +
    geom_bar(position="dodge",stat="identity") + 
    geom_text(aes(label=Percentage), hjust=1, color="white", size=3.5) +
    theme(axis.title.y=element_blank(),
      axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
       legend.title = element_blank()) +
    coord_flip()
    print(g1)
```

Major Problems

``` {r problems}
table <- summary(Incidents$Problem)
Problem_Highest_Freq = data.frame(table[table > 3000])
colnames(Problem_Highest_Freq) <- "Count"
Problem_Highest_Freq$Problem_Type = row.names(Problem_Highest_Freq)

colors <- c("Altered Mentation" = "#6930c3", 
            "Attended Patient" = "#6930c3", 
            "Chest Pain" = "#6930c3", 
            "Community Health Assist" = "#6930c3", 
            "Fall" = "#6930c3", 
            "Hemorrhage" = "#6930c3", 
            "Respiratory" = "#6930c3", 
            "Seizure" = "#6930c3", 
            "Sick" = "#6930c3", 
            "Syncopal Episode" = "#6930c3", 
            "Traffic Injury" = "#6930c3", 
            "Unconscious" = "#6930c3",
            "Pandemic" = "#ef233c",
            "Stroke" = "#8d99ae",
            "Abdominal Pain" = "#8d99ae",
            "Alarm Activation" = "#8d99ae",
            "Assault" = "#8d99ae",
            "Injury" = "#8d99ae",
            "Cardiac Arrest" = "#8d99ae",
            "Psychiatric" = "#8d99ae",
            "Overdose" = "#8d99ae", 
            "Diabetic" = "#8d99ae")


ggplot(data=Problem_Highest_Freq, aes(x=reorder(Problem_Type, Count),y=Count, fill = Problem_Type)) +
    scale_fill_manual(limits = c("Overdose", 
            "Diabetic", 
            "Altered Mentation", 
            "Attended Patient", 
            "Chest Pain", 
            "Community Health Assist", 
            "Fall", 
            "Hemorrhage", 
            "Respiratory", 
            "Seizure", 
            "Sick", 
            "Syncopal Episode", 
            "Traffic Injury", 
            "Unconscious",
            "Pandemic",
            "Stroke",
            "Abdominal Pain",
            "Alarm Activation",
            "Assault",
            "Injury",
            "Cardiac Arrest",
            "Psychiatric"), values = colors) +
    geom_bar(position="dodge",stat="identity") + 
    geom_text(aes(label=Count), hjust=1, color="white", size=3.5) +
    theme(axis.title.y=element_blank(),
      axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
       legend.title = element_blank()) +
    coord_flip() + theme(legend.position = "none")

print("Proportion of the major problems")
sum(Problem_Highest_Freq$Count)/nrow(Incidents)
```

Priority

``` {r priority}
ggplot(Incidents, aes(x=as.factor(Priority_Number), fill = Priority_Number)) + 
  geom_bar(stat = "count") + scale_fill_gradient2(midpoint = 6, low="red", mid = "white", high="light blue") +
  geom_text(stat='count', aes(label=..count..), hjust=0.2, size=3.5) + 
  scale_x_discrete(limits = rev) + coord_flip() + 
  xlab("Priority Number") + ylab("Count") + theme(legend.position = "none")

print("Proportion of high priority")
mean(Incidents$Priority_Number <= 2)
print("Proportion of mid priority")
mean(Incidents$Priority_Number == 3 | Incidents$Priority_Number == 4)
print("Proportion of low priority")
mean(Incidents$Priority_Number >= 5)
```

Response time histogram

``` {r Service Time Distribution}
Incidents_assigned <- Incidents[Incidents$Time_First_Unit_Assigned >= 0 & !is.na(Incidents$Time_First_Unit_Assigned),]$Time_First_Unit_Assigned

Incidents_Enroute <- Incidents[Incidents$Time_First_Unit_Enroute >= 0 &!is.na(Incidents$Time_First_Unit_Enroute),]$Time_First_Unit_Enroute

Incidents_Arrived <- Incidents[Incidents$Time_First_Unit_Arrived >= 0 & !is.na(Incidents$Time_First_Unit_Arrived),]$Time_First_Unit_Arrived

par(mfrow=c(3,1))

length(Incidents_Enroute)

hist(Incidents_assigned, xlim = c(0,5), breaks = max(Incidents_assigned)*20, xlab = "Assignment Time (minutes)", main = "")
legend("topright", legend=c("mean", "median"),
       col=c("blue", "purple"), lty = 1:2, cex=0.8)
abline(v=mean(Incidents_assigned), col="purple", lwd=2)
mean(Incidents_assigned)
abline(v=median(Incidents_assigned), col="blue", lwd=2, lty = 2)
median(Incidents_assigned)

hist(Incidents_Enroute, xlim = c(0,5), breaks = max(Incidents_Enroute)*20, xlab = "Dispatch Time (minutes)", main = "")
abline(v=mean(Incidents_Enroute), col="purple", lwd=2)
mean(Incidents_Enroute)
abline(v=median(Incidents_Enroute), col="blue", lwd=2, lty = 2)
median(Incidents_Enroute)

hist(Incidents_Arrived, xlim = c(0,20), breaks = max(Incidents_Arrived)*20, xlab = "Arrival Time (minutes)", main = "")
abline(v=mean(Incidents_Arrived), col="purple", lwd=2)
mean(Incidents_Arrived)
abline(v=median(Incidents_Arrived), col="blue", lwd=2, lty = 2)
median(Incidents_Arrived)
```

Major hospital response time anova analysis.

```{r anova analysis hospital service time}
library(gplots)
library(dplyr)

print("Assignment Time")

Incidents_assigned <- Incidents[Incidents$Time_First_Unit_Assigned >= 0 & !is.na(Incidents$Time_First_Unit_Assigned),]

Incidents_assigned_Major_hospital <- Incidents_assigned[Incidents_assigned$Call_Disposition %in% c("Dell Seton Med Ctr", "South Austin Hospital", "Saint Davids Med Ctr", "North Austin Hospital", "Seton Med Ctr", "Seton Northwest"),]

ggplot(Incidents_assigned_Major_hospital, aes(x=Call_Disposition, y=Time_First_Unit_Assigned)) +
  geom_violin(trim=FALSE) + stat_summary(fun=mean, geom="point", size=2, color="red") + 
  xlab("") + ylab("Assignment Time") + coord_flip() + ylim(0,5)

Incidents_assigned_Major_hospital %>% group_by(Call_Disposition) %>% summarise(Mean=mean(Time_First_Unit_Assigned), Median=median(Time_First_Unit_Assigned))

fit <- aov(Time_First_Unit_Assigned ~ Call_Disposition, data = Incidents_assigned_Major_hospital)
summary(fit)

print("Dispatch Time")

Incidents_Enroute <- Incidents[Incidents$Time_First_Unit_Enroute >= 0 &!is.na(Incidents$Time_First_Unit_Enroute),]

Incidents_Enroute_Major_hospital <- Incidents_Enroute[Incidents_Enroute$Call_Disposition %in% c("Dell Seton Med Ctr", "South Austin Hospital", "Saint Davids Med Ctr", "North Austin Hospital", "Seton Med Ctr", "Seton Northwest"),]

ggplot(Incidents_Enroute_Major_hospital, aes(x=Call_Disposition, y=Time_First_Unit_Enroute)) +
  geom_violin(trim=FALSE) + stat_summary(fun=mean, geom="point", size=2, color="red") +
  xlab("") + ylab("Dispatch Time") + coord_flip() + ylim(0,5)

Incidents_Enroute_Major_hospital %>% group_by(Call_Disposition) %>% summarise(Mean=mean(Time_First_Unit_Enroute), Median=median(Time_First_Unit_Enroute))

fit <- aov(Time_First_Unit_Enroute ~ Call_Disposition, data = Incidents_Enroute_Major_hospital)
summary(fit)

print("Arrival Time")

Incidents_Arrived <- Incidents[Incidents$Time_First_Unit_Arrived >= 0 & !is.na(Incidents$Time_First_Unit_Arrived),]

Incidents_Arrived_Major_hospital <- Incidents_Arrived[Incidents_Arrived$Call_Disposition %in% c("Dell Seton Med Ctr", "South Austin Hospital", "Saint Davids Med Ctr", "North Austin Hospital", "Seton Med Ctr", "Seton Northwest"),]

ggplot(Incidents_Arrived_Major_hospital, aes(x=Call_Disposition, y=Time_First_Unit_Arrived)) +
  geom_violin(trim=FALSE) + stat_summary(fun=mean, geom="point", size=2, color="red") + 
  xlab("") + ylab("Arrival Time") + coord_flip() + ylim(0,30)

Incidents_Arrived_Major_hospital %>% group_by(Call_Disposition) %>% summarise(Mean=mean(Time_First_Unit_Arrived), Median=median(Time_First_Unit_Arrived))

fit <- aov(Time_First_Unit_Arrived ~ Call_Disposition, data = Incidents_Arrived_Major_hospital)
summary(fit)

plotmeans(Time_First_Unit_Assigned ~ Call_Disposition, data = Incidents_assigned_Major_hospital, xlab="Hospital", cex.axis = 0.5, ylab="Assignment Time(minutes)", 
             main="Mean Plot with 95% CI")
plotmeans(Time_First_Unit_Enroute ~ Call_Disposition, data = Incidents_Enroute_Major_hospital, xlab="Hospital", cex.axis = 0.5, ylab="Dispatch Time(minutes)", 
             main="Mean Plot with 95% CI")
plotmeans(Time_First_Unit_Arrived ~ Call_Disposition, data = Incidents_Arrived_Major_hospital, xlab="Hospital", cex.axis = 0.5, ylab="Arrival Time(minutes)", 
             main="Mean Plot with 95% CI")
```


## Time Series Data Processing

to_daily_ts is a function that converts incidents data into daily counts. 

``` {r ts}
library(stats)
library(forecast)
library(car)
library(astsa)
library(nlme)
library(gridExtra)
library(TSA)
library(xts)
library(ggplot2)
library(lubridate)
library(tseries)
last_day <- as.Date("2020-12-31")
first_day <- as.Date("2019/01/01")
```

Time series of overall EMS calls.

``` {r total ts}
total_ts <- ts(ts.df$total_ts, start = first_day)
summary(total_ts)
sd(total_ts)
```

Time series of non-pandemic defunct EMS calls and pandemic defunct EMS calls.

``` {r defunct ts}
defunct_calls_pandemic_removed_ts <- ts(ts.df$defunct_calls_pandemic_removed_ts, start = first_day)
Pandemic_defunct_calls_ts <- ts(ts.df$Pandemic_defunct_calls_ts, start = first_day)
```

Time series of non-pandemic EMS calls and pandemic EMS calls.

``` {r effect ts}
Incidents_pandemic_removed_ts <- ts(ts.df$Incidents_pandemic_removed_ts, start = first_day)
Pandemic_effect_ts <- ts(ts.df$Pandemic_effect_ts, start = first_day)
```

Time series of daily Covid-19 hospitalization data.

``` {r hospitalisation_ts}
hospitalisation_ts <- ts(ts.df$hospitalisation_ts, start = first_day)
```

Create smoothed time series.

``` {r ts smooth}
library(changepoint.np)
library(tseries)
library(MTS)
library(ppcor)

last_day_int <- as.numeric(last_day - first_day) + 1
first_day_int <- 7

# start is the start day of the smoothing, it is also used as an averaging denominator
# end is the last day of the data
smooth_ts <- function(data, start, end){
  smoothed_data_ts <- c()
  for (i in start:end){
    begin = i - start + 1
    smoothed_data_ts <- c(smoothed_data_ts,mean(data[c(begin: (begin + start - 1))]))
  }
  for (j in start - 2:ceiling(start/2)){
    begin = end - j
    smoothed_data_ts <- c(smoothed_data_ts,mean(data[c(begin : end)]))
  }
  return(ts(smoothed_data_ts, start = first_day + floor(start/2)))
}

smoothed_Incidents_pandemic_removed_ts <- smooth_ts(Incidents_pandemic_removed_ts, start = first_day_int, end = last_day_int)
smoothed_Pandemic_effect_ts <- smooth_ts(Pandemic_effect_ts, start = first_day_int, end = last_day_int)

smoothed_defunct_calls_pandemic_removed_ts <- smooth_ts(defunct_calls_pandemic_removed_ts, start = first_day_int, end = last_day_int)
smoothed_Pandemic_defunct_calls_ts <- smooth_ts(Pandemic_defunct_calls_ts, start = first_day_int, end = last_day_int)

smoothed_hospitalisation_ts <- smooth_ts(hospitalisation_ts, start = first_day_int, end = last_day_int)
```

## Impact of the pandemic

Change point detection on non-pandemic incident time series.

```{r cp_detection comparison}
tt <- time(Incidents_pandemic_removed_ts)
fmt <- "%m-%d-%y" # format for axis labels
raw.man=cpt.meanvar(Incidents_pandemic_removed_ts,method='BinSeg',penalty='BIC',test.stat='Normal', Q = 2)
first_day + cpts(raw.man) - 1

## Comparing with other penalty types and methods
for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.meanvar(Incidents_pandemic_removed_ts,method='BinSeg',penalty=penalty_,
                       pen.value=c(log(length(Incidents_pandemic_removed_ts)),100*log(length(Incidents_pandemic_removed_ts))), test.stat='Normal', Q = 10)
  print(first_day + cpts(raw.man1) - 1)
}

for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.meanvar(Incidents_pandemic_removed_ts,method='PELT',penalty=penalty_, test.stat='Normal')

  print(first_day + cpts(raw.man1) - 1)
}

for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.var(Incidents_pandemic_removed_ts,method='PELT',penalty=penalty_, test.stat='Normal')

  print(first_day + cpts(raw.man1) - 1)
}

```

Change point statistics and time series plot

```{r cp_detection stats and plot}
param.est(raw.man)
print("Standard Deviation")
sqrt(param.est(raw.man)$variance)
print("Change point Dates")
first_day + cpts(raw.man) - 1
labs <- format(as.Date(tt[cpts(raw.man)]), fmt)
plot(raw.man,cpt.width=3, ylab = "Daily Frequency", xaxt ="n", main = "Non-Pandemic EMS Incidents", xlab = "")
axis(side = 1, at = tt[cpts(raw.man)], labels = labs,  cex.axis = 0.5)
param.est(raw.man)$mean[3]/param.est(raw.man)$mean[1]


plot(Incidents_pandemic_removed_ts, xaxt ="n", ylim = c(0, 350), ylab = "Daily Frequency", 
     col = alpha("black", 0.2))
lines(smoothed_Incidents_pandemic_removed_ts, col = "black")
lines(defunct_calls_pandemic_removed_ts,col = alpha("orange", 0.2))
lines(smoothed_defunct_calls_pandemic_removed_ts, col = "orange")
lines(Pandemic_effect_ts,col = alpha("red", 0.2))
lines(smoothed_Pandemic_effect_ts, col = "red")
lines(Pandemic_defunct_calls_ts,col = alpha("brown", 0.2))
lines(smoothed_Pandemic_defunct_calls_ts, col = "brown")
lines(hospitalisation_ts, col = alpha("green", 0.2))
lines(smoothed_hospitalisation_ts, col = "green")
tt <- time(Incidents_pandemic_removed_ts)
for (i in tt[cpts(raw.man)]){
  abline(v=i, col="purple", lwd=2)
}
ix <- seq(1, length(tt), by=50)
fmt <- "%m-%d-%y" # format for axis labels
labs <- format(as.Date(tt[cpts(raw.man)]), fmt)
axis(side = 1, at = tt[cpts(raw.man)], labels = labs, cex.axis = 0.5)
legend("topleft", legend=c("Pandemic Effect", "Pandemic Defunct", "Non-pandemic Defunct"),
       col=c("red", "brown","orange"),lty=1, cex=0.8)
legend("topright", legend=c("Hospitalization", "Non-pandemic Effect"),
       col=c("green", "black"),lty=1, cex=0.8)
```

Comparison of number of Non-Pandemic Defunct EMS incidents per day among Period 1 (before March 17th), Period 2 (March 18th - May 12th), Period 3 (after May 13th). (Table 4 in the paper.)

``` {r }
print("Change points")
cpts(raw.man)

print("n of period 2 and 3")
length(defunct_calls_pandemic_removed_ts[(cpts(raw.man)[1] + 1):(cpts(raw.man)[2] - 1)])
length(defunct_calls_pandemic_removed_ts[cpts(raw.man)[2]:last_day_int])

print("Mean: period 1, 2 and 3")
mean(defunct_calls_pandemic_removed_ts[1:cpts(raw.man)[1]])
mean(defunct_calls_pandemic_removed_ts[(cpts(raw.man)[1] + 1):(cpts(raw.man)[2] - 1)])
mean(defunct_calls_pandemic_removed_ts[cpts(raw.man)[2]:last_day_int])

print("sd: period 1, 2 and 3")
sd(defunct_calls_pandemic_removed_ts[1:cpts(raw.man)[1]])
sd(defunct_calls_pandemic_removed_ts[(cpts(raw.man)[1] + 1):(cpts(raw.man)[2] - 1)])
sd(defunct_calls_pandemic_removed_ts[cpts(raw.man)[2]:last_day_int])

print("Proportion of nonpandemic defunct calls pre (period 1) and post covid(period 3)")
sum(defunct_calls_pandemic_removed_ts[1:442]) / sum(Incidents_pandemic_removed_ts[1:442] + defunct_calls_pandemic_removed_ts[1:442])
sum(defunct_calls_pandemic_removed_ts[499:last_day_int]) / sum(Incidents_pandemic_removed_ts[499:last_day_int] + defunct_calls_pandemic_removed_ts[499:last_day_int])

mean(defunct_calls_pandemic_removed_ts[1:442] / (Incidents_pandemic_removed_ts[1:442] + defunct_calls_pandemic_removed_ts[1:442]))
mean(defunct_calls_pandemic_removed_ts[499:last_day_int] / (Incidents_pandemic_removed_ts[499:last_day_int] + defunct_calls_pandemic_removed_ts[499:last_day_int]))
```

Major problems impacted by the pandemic.

``` {r problems impacted by the pandemic}
problem_ts <- function(data){
  to_daily_ts <- c()
  for (i in first_day:last_day){
    subset <- as.Date(data$Time_PhonePickUp_Date) == as.Date(i)
    data = data[!subset,]
    to_daily_ts <- c(to_daily_ts, sum(subset))
  }
  return(to_daily_ts)
}

problem_TS_Plot <- function(Problem_Type){
  
  Problem_DF <- data.frame()
  
  for (problem in Problem_Type){
    data <- Incidents[Incidents$Problem %in% problem,]
    df <- data.frame(problem_ts(data))
    colnames(df) <- "Daily_Count"
    df$Problem_Type <- problem
    Problem_DF <- rbind(Problem_DF, df)
  }

  return(Problem_DF)
}

Problem_DF <- problem_TS_Plot(Problem_Highest_Freq$Problem_Type)

for (problem in Problem_Highest_Freq$Problem_Type){
  data <- Problem_DF[Problem_DF$Problem_Type == problem,]$Daily_Count
  before <- data[1:cpts(raw.man)[1]]
  mid <- data[(cpts(raw.man)[1] + 1): (cpts(raw.man)[2] - 1)]
  after <- data[cpts(raw.man)[2]:last_day_int]
  print(problem)
  print(t.test(before, after, alternative = "greater"))
  print(mean(mid))
}
```

## Response Time Analysis

Impact of the pandemic on average response time.

``` {r Service Time Distribution COVID}
pre_mid_post_service_time <- function(condition, data, maxtime, title){
  precovid <- data[condition$Time_PhonePickUp_Date <= as.Date("2020-03-17")]

  midcovid <- data[condition$Time_PhonePickUp_Date > as.Date("2020-03-17") & condition$Time_PhonePickUp_Date < as.Date("2020-05-13")]

  postcovid <- data[condition$Time_PhonePickUp_Date >= as.Date("2020-05-13")]
  
  print("Period 1")
  print(summary(precovid))
  print("Period 2")
  print(summary(midcovid))
  print("Period 3")
  print(summary(postcovid))

hist(precovid, xlim = c(0,maxtime), breaks = max(precovid)*20, xlab = "Period 1", main = title)
abline(v=mean(precovid), col="purple", lwd=2)
abline(v=median(precovid), col="blue", lwd=2, lty = 2)

hist(midcovid, xlim = c(0,maxtime), breaks = max(midcovid)*20, xlab = "Period 2", main = "")
abline(v=mean(midcovid), col="purple", lwd=2)
abline(v=median(midcovid), col="blue", lwd=2, lty = 2)

print(t.test(precovid, midcovid, alternative = "less"))

hist(postcovid, xlim = c(0,maxtime), breaks = max(postcovid)*20, xlab = "Period 3", main = "")
abline(v=mean(postcovid), col="purple", lwd=2)
abline(v=median(postcovid), col="blue", lwd=2, lty = 2)

print(t.test(precovid, postcovid, alternative = "less"))
}

par(mfcol=c(3,3))

Incidents_assigned <- Incidents[Incidents$Time_First_Unit_Assigned >= 0 & !is.na(Incidents$Time_First_Unit_Assigned),]
Incidents_assigned$Time_PhonePickUp_Date <- as.Date(Incidents_assigned$Time_PhonePickUp_Date)

pre_mid_post_service_time(Incidents_assigned, Incidents_assigned$Time_First_Unit_Assigned, 4, "Assignment Time (minutes)")

Incidents_Enroute <- Incidents[Incidents$Time_First_Unit_Enroute >= 0 &!is.na(Incidents$Time_First_Unit_Enroute),]
Incidents_Enroute$Time_PhonePickUp_Date <- as.Date(Incidents_Enroute$Time_PhonePickUp_Date)

pre_mid_post_service_time(Incidents_Enroute, Incidents_Enroute$Time_First_Unit_Enroute, 4, "Dispatch Time (minutes)")

Incidents_Arrived <- Incidents[Incidents$Time_First_Unit_Arrived >= 0 & !is.na(Incidents$Time_First_Unit_Arrived),]
Incidents_Arrived$Time_PhonePickUp_Date <- as.Date(Incidents_Arrived$Time_PhonePickUp_Date)
pre_mid_post_service_time(Incidents_Arrived, Incidents_Arrived$Time_First_Unit_Arrived, 15, "Arrival Time (minutes)")
```


Impact of the pandemic on response time for high priority incidents.

``` {r Service Time Analysis - High Priority}

high_priority_proportion <- function(data){
  overall_len = nrow(data)
  data = data[data$Priority_Number <= 2,]
  print("Frequncy of High Priority Incidents:")
  print(nrow(data))
  print("Frequncy of High Priority Incidents with less than 5 min response time:")
  print(sum(data$Time_First_Unit_Arrived <= 5))
  return(data$Time_First_Unit_Arrived <= 5)
}

pre_mid_post_high_priority_service_time <- function(data){
  print("Period 1")
  precovid <- high_priority_proportion(data[data$Time_PhonePickUp_Date <= as.Date("2020-03-17"),])
  print("Proportion of High Priority Incidents whose response time is less than 5 min:")
  print(mean(precovid))
  
  print("Period 2")
  midcovid <- high_priority_proportion(data[data$Time_PhonePickUp_Date > as.Date("2020-03-17") & data$Time_PhonePickUp_Date < as.Date("2020-05-13"),])
  print("Proportion of High Priority Incidents whose response time is less than 5 min:")
  print(mean(midcovid))

  print("Period 3")
  postcovid <- high_priority_proportion(data[data$Time_PhonePickUp_Date >= as.Date("2020-05-13"),])
  
  print("Proportion of High Priority Incidents whose response time is less than 5 min:")
  print(mean(postcovid))

  print(t.test(precovid, postcovid, alternative = "greater"))
}

pre_mid_post_high_priority_service_time(Incidents_Arrived)

```


## Predictive Model Training

Hospitalisation data change point detection.

```{r hospitalisation cp}
raw.man=cpt.var(hospitalisation_ts,method='PELT',penalty='MBIC',test.stat='Normal')
first_day + cpts(raw.man) - 1 

## Comparing with other penalty types and methods
for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.var(hospitalisation_ts,method='PELT', penalty=penalty_,test.stat='Normal')
  print(first_day + cpts(raw.man1) - 1)
}

for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.meanvar(hospitalisation_ts,method='BinSeg',penalty=penalty_, pen.value=c(log(length(hospitalisation_ts)),100*log(length(hospitalisation_ts))), test.stat='Normal', Q = 10)
  print(first_day + cpts(raw.man1) - 1)
}

for (penalty_ in c('BIC', 'SIC', 'MBIC', 'AIC')){
  raw.man1=cpt.var(hospitalisation_ts,method='BinSeg',penalty=penalty_, pen.value=c(log(length(hospitalisation_ts)),100*log(length(hospitalisation_ts))), test.stat='Normal', Q = 10)
  print(first_day + cpts(raw.man1) - 1)
}
```

Plots and statistics of hospitalisation data change point detection.

``` {r hospitalisation cp plots}
tt <- time(hospitalisation_ts)
ix <- seq(1, length(tt), by=50)
fmt <- "%m-%d-%y" # format for axis labels
labs <- format(as.Date(tt[ix]), fmt)

plot(smoothed_hospitalisation_ts, col = "green", ylab = "Daily Count", xaxt ="n", main = "Hospitalization", xlab = "", xlim = first_day + c(420, last_day_int), ylim = c(0, 100))
lines(ts(hospitalisation_ts[1:last_day_int], start = first_day), col = alpha("green", 0.3))
axis(side = 1, at = tt[ix], labels = labs,  cex.axis = 0.5)
lines(Pandemic_effect_ts,col = alpha("red", 0.2))
lines(smoothed_Pandemic_effect_ts, col = "red")
lines(Pandemic_defunct_calls_ts,col = alpha("brown", 0.2))
lines(smoothed_Pandemic_defunct_calls_ts, col = "brown")
legend("topleft", legend=c("Pandemic Effect", "Pandemic Defunct", "Hospitalization"),
       col=c("black", "brown", "green"),lty=1, cex=0.8)

print("Change point Dates")
first_day + cpts(raw.man) - 1 
param.est(raw.man)
labs <- format(as.Date(tt[cpts(raw.man)]), fmt)

plot(smoothed_hospitalisation_ts, col = "green", ylab = "Daily Count", xaxt ="n", main = "Hospitalization", xlab = "", xlim = first_day + c(420, last_day_int), ylim = c(0, 100))
lines(ts(hospitalisation_ts[1:last_day_int], start = first_day), col = alpha("green", 0.3))
axis(side = 1, at = tt[cpts(raw.man)], labels = labs,  cex.axis = 0.5)
#lines(Pandemic_effect_ts,col = alpha("red", 0.2))
#lines(smooth_ts(Pandemic_effect_ts, start = 7, end = last_day_int), col = "red")
#lines(Pandemic_defunct_calls_ts,col = alpha("brown", 0.2))
#lines(smooth_ts(Pandemic_defunct_calls_ts, start = 7, end = last_day_int), col = "brown")
for (i in tt[cpts(raw.man)]){
  abline(v=i, col="purple", lwd=2)
}
legend("topleft", legend=c("Hospitalization"),
       col=c("green"),lty=1, cex=0.8)
```

Add dummy variables to incorporate change points.

``` {r plain arima data processing}
smoothed_last_idx <- length(smoothed_hospitalisation_ts)

smoothed_hospitalisation_df <- data.frame(c(smoothed_hospitalisation_ts[(465 - floor(7/2)):smoothed_last_idx]), 
                                          c(rep(1, cpts(raw.man)[2] - (464 - floor(7/2))), rep(0, smoothed_last_idx - cpts(raw.man)[2])), 
                                          c(rep(0, cpts(raw.man)[2] - (464 - floor(7/2))), rep(1, cpts(raw.man)[3] - cpts(raw.man)[2]), rep(0, smoothed_last_idx- cpts(raw.man)[3])), 
                                          c(rep(0, cpts(raw.man)[3] - (464 - floor(7/2))), rep(1, cpts(raw.man)[4] - cpts(raw.man)[3]), rep(0, smoothed_last_idx - cpts(raw.man)[4])))
colnames(smoothed_hospitalisation_df) <- c("hosp_ts", "first", "second", "third")
smoothed_hospitalisation_df <- as.matrix(smoothed_hospitalisation_df)

MDF <- data.frame(Pandemic_effect_ts[465:last_day_int], smoothed_hospitalisation_df)
colnames(MDF) <- c("Pandemic", "hosp_ts", "first", "second", "third")
```

Model Fitting: with and without change points.

``` {r model fitting}
library(forecast)
xreg <- as.matrix(MDF[,c("hosp_ts", "first", "second", "third")])
auto.arima(MDF$Pandemic, xreg = xreg, approximation = FALSE, stepwise = FALSE)

library(caTools)
set.seed(1)
data1= sample.split(MDF,SplitRatio = 0.8)
train = subset(MDF,data1==TRUE)
test = subset(MDF,data1==FALSE)

print("With change points")
fit.with_cp <- lm(Pandemic ~ hosp_ts + first + second + third, data = train)
summary(fit.with_cp)
plot(fit.with_cp)

print("MSE")
mean((predict(fit.with_cp, newdata = test)-test$Pandemic)^2)
print("Test Residual standard error")
sd(predict(fit.with_cp, newdata = test)-test$Pandemic)

print("Without change points")
fit.without_cp <- lm(Pandemic ~ hosp_ts, data = train)
summary(fit.without_cp)
plot(fit.without_cp)
print("MSE")
mean((predict(fit.without_cp, newdata = test)-test$Pandemic)^2)
print("Test Residual standard error")
sd(predict(fit.without_cp, newdata = test)-test$Pandemic)
```

Plot prediction interval.

``` {r }
y <- data.frame(predict(fit.with_cp, newdata = MDF, interval = "prediction"))
x <- 1:nrow(y)
plot(y$fit, type = "l", xaxt ="n", ylim = c(0, 70), lwd = 2, col = "blue", ylab = "Daily Frequency", xlab = "")
lines(MDF$Pandemic, type = "l", col = alpha("red"))
axis(side = 1, at = cpts(raw.man) - 465, labels = labs,  cex.axis = 1)
legend("topright", legend=c("predicted", "actual"),
       col=c("blue", "red"),lty=1, cex=0.8)
  CI.x.top <- rev(x) # x values going forward
  CI.x.bot <- x # x values backwards
  CI.x <- c(CI.x.top, CI.x.bot)

  CI.y.top <- rev(y$upr) # x values going forward
  CI.y.bot <- y$lwr # x values backwards
  CI.y <- c(CI.y.top, CI.y.bot)

  CI.col <- adjustcolor("blue",alpha.f=0.25) # Pick a pretty CI color
  polygon(CI.x, CI.y, col=CI.col, border=NA)
```

Obtain $r^2$ valude on smoothed data

``` {r r squared}
temp_data = data.frame(smoothed_Pandemic_effect_ts[462:(462 + nrow(MDF)-1)], smoothed_hospitalisation_df[1:nrow(MDF),])
colnames(temp_data) <- c("Pandemic", "hosp_ts", "first", "second", "third")
train = subset(temp_data,data1==TRUE)
test = subset(temp_data,data1==FALSE)

get_rsq <- function(data, model){
  rss <- sum((predict(model, newdata = data) - data$Pandemic) ^ 2)  ## residual sum of squares
  tss <- sum((data$Pandemic - mean(data$Pandemic)) ^ 2)  ## total sum of squares
  rsq <- 1 - rss/tss
  return(rsq)
}

print("With changepoint")
print("train R squared")
get_rsq(train, fit.with_cp)
print("test R squared")
get_rsq(test, fit.with_cp)

print("Without changepoint")
print("train R squared")
get_rsq(train, fit.without_cp)
print("test R squared")
get_rsq(test, fit.without_cp)
```